A Mechanised Library for Generic (Co)Programming

### 2024-02-22 paper/paper.tex
Working on background section 2.

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\dcas}[1]{\textcolor{red}{\textsc{David}: #1}}
\newcommand{\mvol}[1]{\textcolor{red}{\textsc{Michael}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}
\newcommand{\curry}{\Lambda}
\newcommand{\uncurry}{\begin{sideways}\begin{sideways}$\Lambda$\end{sideways}\end{sideways}}
\newcommand{\Bool}{\mathbb{B}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Nat}{\N}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Sets}{\mathbf{Set}}
\newcommand{\blacklater}{\blacktriangleright}
\newcommand{\tot}{\mathcal{S}}
\newcommand{\PSh}{\ensuremath{\textbf{PSh}(\omega)}}

@@

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}

\title{A Mechanised Library for Generic (Co)Programming}

@@

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorithm. This is given by the
\haskell{partition} and \haskell{quicksort} function below (written in Haskell
code):
\begin{minted}{haskell}
partition :: [a] -> ([a], a, [a])
partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)

quicksort :: [a] -> [a]
quicksort [] = []
quicksort ls = let (xs, x, ys) = partition ls in
                        (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
While \haskell{quicksort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \haskell{partition}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

@@

\section{Recursive Equations}

@@

\[
  f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
\]

\subsubsection{Example}
Say that we want to prove that the two functions below are well-defined

One option is to prove that the following function (given by a catamorphism) is
equivalent to the above
\begin{align*}
  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
  & \operator{even-odd} (0) = (\top, \bot)\\
  & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
\end{align*}
This function is clearly a catarmorphism:
\begin{align*}
  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
  & \operator{even-odd} = \cata{a} \operator{ where }\\
  & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
  & \qquad \qquad a (\Nil) = (\top, \bot)\\
  & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
\end{align*}
where $\cata{\cdot}$ takes a $(1+-)$ algebra on $\Bool \times \Bool$ and returns
the associated recursive function.

% A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$

@@

\subsection{Course-of-Value Recursion with Finite Memory Table}
It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for
any algebra $a$ generates a finite output.  In other words, as far as the
hylomorphism is concerned $G_{\infty}$ can be regarded as its finite version
\[
  G_{*} A = \mu X. A \times G X
\]
We now rewrite course-of-value recursion using the finite memory table
For an algebra $a : D(D_{*} A) \to A$,
\begin{align*}
  & g : \mu F \to A
  & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
\end{align*}

@@

In Coq, to construct least/greatest fixpoints of functors, these need to build
\emph{strictly positive types}. A common approach to construct such strictly
positive endofunctors is to use \emph{container functors}~\cite{Whatever}.
However, reasoning about container and container functor equality requires to
consider: (a) functional extensionality; (b) heterogeneous equality. This is
problematic, due to our goals to remain axiom-free. Furthermore, standard
representations of containers in Coq will lead to OCaml code that contains
\mintinline{OCaml}{Obj.magic}. We will explain why, and our mechanisation of
containers in the remainder of this section.

@@

\bibliography{references.bib}
\appendix

>>>>

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\dcas}[1]{\textcolor{green}{\textsc{David}: #1}}
\newcommand{\mvol}[1]{\textcolor{blue}{\textsc{Michael}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}
\newcommand{\curry}{\Lambda}
\newcommand{\uncurry}{\begin{sideways}\begin{sideways}$\Lambda$\end{sideways}\end{sideways}}
\newcommand{\Bool}{\mathbb{B}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Nat}{\N}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Sets}{\mathbf{Set}}
\newcommand{\blacklater}{\blacktriangleright}
\newcommand{\tot}{\mathcal{S}}
\newcommand{\PSh}{\ensuremath{\textbf{PSh}(\omega)}}

@@

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\title{A Mechanised Library for Generic (Co)Programming}

@@

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorithm. This is given by the
\haskell{partition} and \haskell{quicksort} function below (written in Haskell
code):
While \haskell{quicksort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \haskell{partition}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

@@

\section{Background on Recursion Schemes}
Recursion schemes provide a systematic way to work with data types, such as
trees or lists, by abstracting away the recursion into higher-order functions.
Canonical recursion schemes include \emph{folds} (catamorphisms) which consume
data and \emph{unfolds} (anamorphisms) which produce data. These are the unique
solutions (when they exist) to the following two equations respectively
\[
  h \comp \InIso = a \comp F h \qquad \OutIso \comp k = F k \comp c
\]
where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
$F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
$F$-coalgebra), $\InIso$ is the isomoprhism on the least fixed-point of $F$,
namely $\mu F \xrightarrow{\iso} F\mu F$, and $\OutIso$ is the isomorphism on
the greatest fixed-point of $F$, namely $\nu F \xrightarrow{\iso} F\nu F$. When
the unique solutions to the above equations exist we denote them as $\cata{a}$
and $\ana{c}$ respectively.

However, catamorphisms (dually anamorphisms) do not capture every recursion
scheme we can think of.  As an example, consider the quicksort
algorithm:\footnote[1]{This program is written in Coq notation to be consistent,
but it is not a program that Coq can accept.}

\mpav{I need this to be written in Coq notation}
\begin{minted}{haskell}
partition :: [a] -> ([a], a, [a])
partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)

quicksort :: [a] -> [a]
quicksort [] = []
quicksort ls = let (xs, x, ys) = partition ls in
                        (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
This program is formed of essentially two parts. The first one is the
$\operator{partition}$ program destructuring the input list in three parts, the
pivot and the other two sublists, and the second part recursing over the
sublists and composing the results back together.

Notice that if we define the functor \coq{TreeF} $X = X \times A \times X$ with
the obvious functorial action then \coq{partition} can be viewed as a
\coq{TreeF}-coalgebra on lists over the type $A$. While we can define an \coq{TreeF}-algebra on lists over $A$
as well, that is a map \coq{alg} $:$ \coq{TreeF} $[A] \to [A]$ defined
as \coq{alg (left, x, right) = left ++ [x] ++ right}.

More specifically, in the context of recursive data types, such as in Haskell,
where induction and coinduction is conflated, an hylomorphism can be implemented
in terms of a catarmorphism (folding) and an anamorphism (unfolding)
\[
  \hylo{a}{c} = \cata{a} \comp \ana{c}
\]
They first generate a data structure using an anamorphism and then immediately
fold or reduce it using a catamorphism.

The hylomorphism pattern is particularly useful for defining recursive
algorithms in a functional programming context, where recursion is a fundamental
tool for solving problems. By separating the generation of data structures from
their consumption, hylomorphisms enable clearer and more modular code, making it
easier to reason about and maintain recursive algorithms.

However, in a type theory inductive and coinductive data types are separate.

In this section we will introduce the notion of hylomorphism. By way of examples
we will also give high level intution as to way all recursion schemes can be
defined in terms of a basic

Say that we want to prove that the two functions below are well-defined

One option is to prove that the following function is
equivalent to the above
\begin{align*}
  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
  & \operator{even-odd} (0) = (\top, \bot)\\
  & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
\end{align*}
This function is clearly a catarmorphism:
\begin{align*}
  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
  & \operator{even-odd} = \cata{a} \operator{ where }\\
  & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
  & \qquad \qquad a (\Nil) = (\top, \bot)\\
  & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
\end{align*}
where $\cata{\cdot}$ takes a $(1+-)$ algebra on $\Bool \times \Bool$ and returns
the associated recursive function.

@@

\[
  f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
\]
% A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$

@@

@@

In Coq, to construct least/greatest fixpoints of functors, these need to build
\emph{strictly positive types}. A common approach to construct such strictly
positive endofunctors is to use \emph{container functors}~\cite{AbbottAG05}.
However, reasoning about container and container functor equality requires to
consider: (a) functional extensionality; (b) heterogeneous equality. This is
problematic, due to our goals to remain axiom-free. Furthermore, standard
representations of containers in Coq will lead to OCaml code that contains
\mintinline{OCaml}{Obj.magic}. We will explain why, and our mechanisation of
containers in the remainder of this section.

@@

\bibliography{references.bib}
\appendix
\subsection{Course-of-Value Recursion with Finite Memory Table}
It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for
any algebra $a$ generates a finite output.  In other words, as far as the
hylomorphism is concerned $G_{\infty}$ can be regarded as its finite version
\[
  G_{*} A = \mu X. A \times G X
\]
We now rewrite course-of-value recursion using the finite memory table
For an algebra $a : D(D_{*} A) \to A$,
\begin{align*}
  & g : \mu F \to A
  & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
\end{align*}
